Fast approximate Bayesian inference of HIV indicators using PCA adaptive Gauss-Hermite quadrature

### 2023-06-26 src/docs_paper/paper.Rmd
Easy Alex fixes

Although the approach of `TMB` is fast, within the empirical Bayes framework hyperparameter uncertainty is not properly accounted for in the latent field posterior.
This consideration motivated us to look for an approach closer to full Bayesian inference, which is also flexible enough to be compatible with the model, as well as fast enough to be run in production by country teams.
To obtain fast, accurate Bayesian inferences for the Naomi model we developed an inference method which extends adaptive Gauss-Hermite quadrature (AGHQ) to handle many hyperparameters.
AGHQ is a quadrature method based on the theory of polynomial interpolation, which is well suited to statistical estimation problems.
@bilodeau2022stochastic prove stochastic convergence rates for Bayesian posterior quantities when the normalising constant is estimated using AGHQ.
It is not computationally feasible to use AGHQ in high dimensions directly, as exponentially many nodes would be required.
Instead, we use principal components analysis (PCA) of the Hessian at the mode to find a smaller number of dimensions which explain most of the variance.
In our application, this results in a grid which has millions of times fewer nodes the corresponding dense grid.
Our method is implemented as an extension of the `aghq` \textsf{R} package [@stringer2021implementing].
As `aghq` is designed to naturally interface with `TMB`, use is simple when provided a \textsf{C++} user template for the log-posterior.

@@

Let $\mathbf{y} = (y^{\theta}_I)$ for $\theta \in \{\rho, \alpha, \kappa, \rho^\text{ANC}, \alpha^\text{ANC}, N^\text{ART}\}$ and $I \in \mathcal{I}$ be a concatenated vector of observations.
Here, we could attempt to write the Naomi model fully in a small number of equations, but likely it'd be difficult.

@@

Latent Gaussian models [LGMs; @rue2009approximate] are three-stage hierarchical models of the form
\begin{align*}
y_i &\sim p(y_i \, | \, \eta_i, \btheta_1), \quad i \in [n]\\
\mu_i &= \mathbb{E}(y_i \, | \, \eta_i) = g(\eta_i), \\
\eta_i &= \beta_0 + \sum_{l = 1}^{p} \beta_j z_{ji} + \sum_{k = 1}^{r} f_k(u_{ki}),
\end{align*}
where $[n] = \{1, \ldots, n\}$.
The response variable is $\y = (y)_{i \in [n]}$ with likelihood $p(\y \, | \, \bmeta, \btheta_1) = \prod_{i = 1}^n p(y_i \, | \, \eta_i, \btheta_1)$, where $\bmeta = (\eta)_{i \in [n]}$.
Each response has conditional mean $\mu_i$ with inverse link function $g: \mathbb{R} \to \mathbb{R}$ such that $\mu_i = g(\eta_i)$.
The vector $\btheta_1 \in \mathbb{R}^{s_1}$, with $s_1$ assumed small, are additional parameters of the likelihood.
The structured additive predictor $\eta_i$ may include an intercept $\beta_0$, linear effects $\beta_j$ of the covariates $z_{ji}$, and unknown functions $f_k(\cdot)$ of the covariates $u_{ki}$.
The parameters $\beta_0$, $\{\beta_j\}$, $\{f_k(\cdot)\}$ are each assigned Gaussian priors.
It is convenient to collect these parameters into a vector $\x \in \mathbb{R}^N$ called the latent field such that $\x \sim \mathcal{N}(0, \bm{Q}(\btheta_2)^{-1})$ where $\btheta_2 \in \mathbb{R}^{s_2}$ are further parameters, again with $s_2$ assumed small.
Let $\btheta = (\btheta_1, \btheta_2) \in \mathbb{R}^s$ with $m = s_1 + s_2$ be all hyperparameters, with prior $p(\btheta)$.

Extended latent Gaussian models [ELGMs; @stringer2022fast] relax the restriction that there is a one-to-one mapping between the mean response $\bmu$ and structured additive predictor $\bmeta$.
Instead, the structured additive predictor is redefined as $\bmeta = (\eta)_{i \in [N_n]}$, where $N_n \in \mathbb{N}$ is a function of $n$, and it is possible that $N_n \neq n$.
Each mean response $\mu_i$ now depends on some subset $\mathcal{J}_i \subseteq [N_n]$ of indices of $\bmeta$, with $\cup_{i = 1}^n \mathcal{J}_i = [N_n]$ and $1 \leq |\mathcal{J}_i| \leq N_n$.
The inverse link function $g(\cdot)$ is redefined for each observation to be a possibly many-to-one mapping $g_i: \mathbb{R}^{|\mathcal{J}_i|} \to \mathbb{R}$, such that $\mu_i = g_i(\bmeta_{\mathcal{J}_i})$.
Importantly, this mapping allows for the presence of non-linearity in the model.
Put together, ELGMs are then of the form
\begin{align*}
y_i &\sim p(y_i \, | \, \bmeta_{\mathcal{J}_i}, \btheta_1), \quad i \in [n] \\
\mu_i &= \mathbb{E}(y_i \, | \, \bmeta_{\mathcal{J}_i}) = g_i(\bmeta_{\mathcal{J}_i}), \\
\eta_j &= \beta_0 + \sum_{l = 1}^{p} \beta_j z_{ji} + \sum_{k = 1}^{r} f_k(u_{ki}), \quad j \in [N_n].
\end{align*}

@@

We fit the simplified Naomi model (Section \ref{sec:naomi}) to data from Malawi using three inferential approaches.
The three approaches were:
1. TMB,
2. PCA-AGHQ, and
3. NUTS: the Hamiltonian Monte Carlo (HMC) algorithm No-U-Turn Sampling (NUTS) using Stan [@carpenter2017stan] via the `tmbstan` package [@monnahan2018no].
The `TMB` C++ user-template used to specify the log-posterior was the same for each.
The dimension of the latent field was $N = 467$ and the dimension of the hyperparameters was $m = 24$.
Settings used for each inferential method are provided in Table \ref{tab:inference-methods}, and, where relevant, discussed further below.
For the deterministic methods, following inference we simulated hyperparameter and latent field samples.
For all methods, we simulated age-sex-district specific HIV prevalence, ART coverage and HIV incidence from the latent field and hyperparameter posteriors.
<!-- Note: eventually this figure should be generated using the algorithm proposed in the paper -->
Example model outputs from TMB are illustrated in Figure \ref{fig:naomi-results}.
The \textsc{R} [@r] code used to produce all results we describe below is available at `github.com/athowes/naomi-aghq`.
We used `orderly` [@orderly] for reproducible research, `ggplot2` for data visualisation [@wickham2016ggplot2] and `rticles` [@allaire2022rticles] for reporting via `rmarkdown` [@allaire2022rmarkdown].

>>>>

Although the approach of `TMB` is fast, within the empirical Bayes framework hyperparameter uncertainty is not properly accounted for in the latent field posterior.
This consideration motivated us to look for an approach closer to full Bayesian inference, which is also flexible enough to be compatible with the model, as well as fast enough to be run in production by country teams.
To obtain fast, accurate Bayesian inferences for the Naomi model we developed an inference method which extends adaptive Gauss-Hermite quadrature (AGHQ) to handle many hyperparameters.
AGHQ is a quadrature method based on the theory of polynomial interpolation, which is well suited to statistical estimation problems.
@bilodeau2022stochastic prove stochastic convergence rates for Bayesian posterior quantities when the normalising constant is estimated using AGHQ.
It is not computationally feasible to use AGHQ in high dimensions directly, as exponentially many nodes would be required.
Instead, we use principal components analysis (PCA) of the Hessian at the mode to find a smaller number of dimensions which explain most of the variance.
For the Naomi model in Malawi, this results in a grid which has millions of times fewer nodes the corresponding dense grid.
Our implementation of the method makes use of the existing Naomi `TMB` template, and is immediately compatible with any model with such a template.

@@

In all, we model the observations $\mathbf{y} = (y^{\theta}_I)$ for $\theta \in \{\rho, \alpha, \kappa, \rho^\text{ANC}, \alpha^\text{ANC}, N^\text{ART}\}$ and $I \in \mathcal{I}$.
Let $\x$ be the latent field, comprised of all intercept effects, age random effects, and spatial random effects in the model.
Let $\btheta$ be hyperparameters, comprised of all standard deviations, and BYM2 proportion parameters.

@@

Latent Gaussian models [LGMs; @rue2009approximate] are three-stage hierarchical models with likelihood
\begin{align*}
y_i &\sim p(y_i \, | \, \eta_i, \btheta_1), \quad i \in [n]\\
\mu_i &= \mathbb{E}(y_i \, | \, \eta_i) = g(\eta_i), \\
\eta_i &= \beta_0 + \sum_{l = 1}^{p} \beta_j z_{ji} + \sum_{k = 1}^{r} f_k(u_{ki}),
\end{align*}
where $[n] = \{1, \ldots, n\}$.
The response variable is $\y = (y)_{i \in [n]}$ with likelihood $p(\y \, | \, \bmeta, \btheta_1) = \prod_{i = 1}^n p(y_i \, | \, \eta_i, \btheta_1)$, where $\bmeta = (\eta)_{i \in [n]}$.
Each response has conditional mean $\mu_i$ with inverse link function $g: \mathbb{R} \to \mathbb{R}$ such that $\mu_i = g(\eta_i)$.
The vector $\btheta_1 \in \mathbb{R}^{s_1}$, with $s_1$ assumed small, are additional parameters of the likelihood.
The structured additive predictor $\eta_i$ may include an intercept $\beta_0$, linear effects $\beta_j$ of the covariates $z_{ji}$, and unknown functions $f_k(\cdot)$ of the covariates $u_{ki}$.
The parameters $\beta_0$, $\{\beta_j\}$, $\{f_k(\cdot)\}$ are each assigned Gaussian priors.
It is convenient to collect these parameters into a vector $\x \in \mathbb{R}^N$ called the latent field such that $\x \sim \mathcal{N}(0, \bm{Q}(\btheta_2)^{-1})$ where $\btheta_2 \in \mathbb{R}^{s_2}$ are further parameters, again with $s_2$ assumed small.
Let $\btheta = (\btheta_1, \btheta_2) \in \mathbb{R}^s$ with $m = s_1 + s_2$ be all hyperparameters, with prior $p(\btheta)$.

Extended latent Gaussian models [ELGMs; @stringer2022fast] relax the restriction that there is a one-to-one mapping between the mean response $\bmu$ and structured additive predictor $\bmeta$.
Instead, the structured additive predictor is redefined as $\bmeta = (\eta)_{i \in [N_n]}$, where $N_n \in \mathbb{N}$ is a function of $n$, and it is possible that $N_n \neq n$.
Each mean response $\mu_i$ now depends on some subset $\mathcal{J}_i \subseteq [N_n]$ of indices of $\bmeta$, with $\cup_{i = 1}^n \mathcal{J}_i = [N_n]$ and $1 \leq |\mathcal{J}_i| \leq N_n$.
The inverse link function $g(\cdot)$ is redefined for each observation to be a possibly many-to-one mapping $g_i: \mathbb{R}^{|\mathcal{J}_i|} \to \mathbb{R}$, such that $\mu_i = g_i(\bmeta_{\mathcal{J}_i})$.
Importantly, this mapping allows for the presence of non-linearity in the model.
Put together, ELGMs are then of the form
\begin{align*}
y_i &\sim p(y_i \, | \, \bmeta_{\mathcal{J}_i}, \btheta_1), \quad i \in [n] \\
\mu_i &= \mathbb{E}(y_i \, | \, \bmeta_{\mathcal{J}_i}) = g_i(\bmeta_{\mathcal{J}_i}), \\
\eta_j &= \beta_0 + \sum_{l = 1}^{p} \beta_j z_{ji} + \sum_{k = 1}^{r} f_k(u_{ki}), \quad j \in [N_n],
\end{align*}
with latent field and hyperparameter priors as in the LGM case.

@@

We fit the simplified Naomi model (Section \ref{sec:naomi}) to data from Malawi using three inferential approaches.
These were:
1. TMB,
2. PCA-AGHQ, building upon the `aghq` package [@stringer2021implementing],
3. NUTS: the Hamiltonian Monte Carlo (HMC) algorithm No-U-Turn Sampling (NUTS) using Stan [@carpenter2017stan] via the `tmbstan` package [@monnahan2018no].
Our goal in this analysis was to determine the accuracy of the approximate methods TMB and PCA-AGHQ as compared with the gold-standard NUTS.
The `TMB` C++ user-template used to specify the log-posterior was the same for each approach.
Settings used for each inferential method are provided in Table \ref{tab:inference-methods}, and, where relevant, discussed further below.
The dimension of the latent field was $N = 467$ and the dimension of the hyperparameters was $m = 24$.
For the deterministic methods, following inference we simulated hyperparameter and latent field samples.
For all methods, we simulated age-sex-district specific HIV prevalence, ART coverage and HIV incidence from the latent field and hyperparameter posteriors.
<!-- Note: eventually this figure should be generated using the algorithm proposed in the paper -->
Example model outputs from TMB are illustrated in Figure \ref{fig:naomi-results}.
The \textsc{R} [@r] code used to produce all results we describe below is available at `github.com/athowes/naomi-aghq`.
We used `orderly` [@orderly] for reproducible research, `ggplot2` for data visualisation [@wickham2016ggplot2] and `rticles` [@allaire2022rticles] for reporting via `rmarkdown` [@allaire2022rmarkdown].
